An Experimental Analysis of the Influence of the Precision of Parameters in an Artificial Neural Network

### 2025-07-16 paper/main.tex
added decimal roudning content
fluffed up summary and conclusion

A non training related issue is the storage of the NN parameters. Usually, the parameters
are stored in the format that was chosen to train with. This is best for reproducibility
as well as accuracy when doing inference with the trained model weights. The datatype  
precision has a higher impact during training though, which begs the question if one could
omit precision from the model weights. In this section, we explore a naive quantization
method: Rounding to a fixed number of decimals. With significantly little decimals and a
reasonable range for the integer part, one could store only the significant and provide
base as well as exponent as metadata.

@@

\section{Summary}
\begin{itemize}
	\item \textbf{16-bit is enough for this classification task}: End-to-end float16 preserved the accuracy.
	\item \textbf{Three decimals suffice}: Storing weights down to fractional length $3$ caused no visible accuracy loss.
\end{itemize}

\section{Discussion}

The results of the first experiments have shown that either the selected task
(MNIST classification) was not complex enough to be affected by the decrease in
accuracy, or the accuracy would need to be decreased even further than float16.

Otherwise, the results generally indicate that networks trained and evaluated
with as low as 16-bit precision can achieve comparable performance to their
high-precision counterparts. Moreover, rounding pretrained weights to just 3
decimal digits preserves full accuracy, opening the door to aggressive
post-training quantization.

This is particularly relevant for edge devices and real-time applications where
memory and compute constraints are more relevant.

\section{Conclusion and Future Directions}

This work confirms that for relatively simple tasks like MNIST classification,
significant reductions in parameter precision can be achieved without
compromising performance. Our future efforts will extend this analysis to more
complex datasets (e.g., CIFAR-10, ImageNet) and deeper architectures, and
explore advanced quantization-aware training techniques.

>>>>

A non training related issue is the storage of the NN parameters. Usually, the parameters
are stored in the format that was chosen to train with. This is best for reproducibility
as well as accuracy when doing inference with the trained model weights. The datatype  
precision has a higher impact during training though, which begs the question if one could
omit precision from the model weights. In this section, we explore a naive quantization
method: Rounding to a fixed number of decimals. With significantly little decimals and a
reasonable range for the integer part, one could store only the significant and provide
base as well as exponent as metadata. When loading the model, the weights are converted
back to floating point, with whatever precision is desired. Only the weights are quantized;
the input data, as well are the values propagated through the network are at full
precision.

@@

\section{Summary}
% \begin{itemize}
% 	\item \textbf{16-bit is enough for this classification task}: End-to-end float16 preserved the accuracy.
% 	\item \textbf{Three decimals suffice}: Storing weights down to fractional length $3$ caused no visible accuracy loss.
% \end{itemize}

% \section{Discussion}
% The results of the first experiments have shown that either the selected task (MNIST
% classification) was not complex enough to be affected by the decrease in accuracy, or the
% accuracy would need to be decreased even further than float16. Otherwise, the results
% generally indicate that networks trained and evaluated with as low as 16-bit precision can
% achieve comparable performance to their high-precision counterparts. Moreover, rounding
% pretrained weights to just 3 decimal digits preserves full accuracy, opening the door to
% aggressive post-training quantization.This is particularly relevant for edge devices and
% real-time applications where memory and compute constraints are more relevant.

The experimental results presented in this paper demonstrate the robustness of neural
networks to reduced parameter precision, particularly for simpler tasks like MNIST
classification. Key findings include:

\begin{itemize}
	\item \textbf{Precision Reduction During Training and Inference}: Training and evaluating
	models with 16-bit floating-point (FP16) precision preserved accuracy comparable to
	higher-precision formats (FP32 and FP64). This suggests that for tasks of similar
	complexity, FP16 is sufficient, offering potential memory and computational savings
	without sacrificing performance.

	\item \textbf{Post-Training Quantization}: Reducing precision only during inference, after
	training with high-precision (FP64), maintained baseline accuracy. This is
	particularly valuable for deployment scenarios where training is done on powerful
	hardware, but inference occurs on resource-constrained devices.

	\item \textbf{Mixed-Precision:} Training and Gradient Clipping: TODO

	\item \textbf{Weight Rounding:} Rounding trained weights to just three decimal digits
	(approximately 11 bits) achieved full accuracy, highlighting the potential for
	aggressive quantization in storage-efficient model deployment. This finding is
	significant for edge applications where memory footprint is critical.
\end{itemize}

However, the simplicity of the MNIST dataset and the small network architecture used in
the experiments may limit the generalizability of these findings. More complex tasks
(e.g., CIFAR-10, ImageNet) and deeper architectures (e.g., CNNs, Transformers) could
exhibit greater sensitivity to precision reduction. Additionally, the impact of precision
on training dynamics, such as gradient flow and convergence, was not deeply explored and
could be a focus for future work.

\section{Conclusion}

% This work confirms that for relatively simple tasks like MNIST classification,
% significant reductions in parameter precision can be achieved without
% compromising performance. Our future efforts will extend this analysis to more
% complex datasets (e.g., CIFAR-10, ImageNet) and deeper architectures, and
% explore advanced quantization-aware training techniques.

This study demonstrates that neural networks can maintain strong performance even under
significant reductions in parameter precision, particularly for simpler tasks like MNIST
classification. Key findings confirm that FP16 training and inference, post-training
quantization, and aggressive weight rounding (down to three decimal digits) preserve model
accuracy while improving memory and computational efficiency. These results are promising
for deploying models in resource-constrained environments, such as edge devices and
real-time applications.

To build on these insights and expand their applicability, future work should focus on the
following directions:

\begin{itemize}
\item \textbf{Complex Architectures and Datasets:} Future research should validate these
findings on more complex models (e.g., CNNs, Transformers) and datasets (e.g., CIFAR-10,
ImageNet) to assess the scalability of low-precision techniques.

\item \textbf{Quantization-Aware Training:} Investigating advanced methods like learned
quantization, non-uniform bit allocation, and dynamic precision adjustment could further
optimize the trade-off between efficiency and accuracy.

\item \textbf{Hardware-Specific Optimizations:} Exploring how reduced precision interacts
with specialized hardware (e.g., TPUs, neuromorphic chips) could unlock new efficiency
gains in real-world deployments.

\item \textbf{Theoretical Understanding:} A deeper analysis of how precision affects
gradient dynamics, convergence, and generalization could provide foundational insights for
designing robust low-precision training algorithms.
\end{itemize}
